---
title: "BF_Final_Exam_Fall23"
author: "Daryll Dalmeida"
date: "12/04/2023"
output: html_document
---

```{r}
# Importing Libraries
library(fpp2)
library(readr)

GOLD <- read_csv("GLD_Final.csv")
Gold_Raw <- GOLD$Close
Gold_ts <- ts(Gold_Raw, frequency = 12, start = c(2018,12))
plot(Gold_ts)
```

```{r}
#### 1. Show a time series plot

gold_time_series <- window(Gold_ts, start = c(2018,12))
plot(gold_time_series)

### 2. summarizing observations of the times series plot

# We see that for the time series plot, the gold prices rise tremendously from 2019 to mid 2020 followed by a series of rise and falls after and then eventually rising to the highest.

```



# Central Tendency

# 1. What are the min, max, mean, median, 1st and 3rd Quartile values of the times series?

```{r}
summary(gold_time_series)


# The min value is : 121.2
# The max value is : 188.8
# The Mean value is : 161.9
# The median value is : 166.7
# The 1st and 3rd quartile values are : 149.1 and 177.4 respectively.

```

# 2. box plot

```{r}
boxplot(gold_time_series,  main = "Box Plot of Gold prices Data (from 2018)")
hist(gold_time_series, main = "Histogram Plot of Gold prices Data (from 2018)")
Acf(gold_time_series, main = "Acf of Gold prices Data (from 2018)")
```

# 3. Summarizing observations about the time series from the summary stats and box plot.

# Observations : From the summary, we can figure out that the median value is more than the mean. 
#                The median values are closer to the 3rd quartile values. From the median being more than the   #                mean, the data is right skewed. From the ACF plot, we can analyze that the first 7 lines are   #                 above the treshold limit which says that there is a trend in the data. After the 14th line,the #                ACF plot goes into a negative value and this says that there is seasonality.


## Decomposition

# 1. Plot the decomposition of the time series

```{r}
stl_decomp <- stl(gold_time_series,s.window ="periodic")
plot(stl_decomp, main = 'Decomposition plot')

```

# 2. Is the times series seasonal? : Yes, the times series has a seasonal component.

# 3. Is the decomposition additive or multiplicative?

```{r}
decom <- decompose(gold_time_series)
decom$type
```

# Yes this is additive. 

# 4. If seasonal, what are the values of the seasonal monthly indices?

```{r}
decom$figure
```

# 5. For which month is the value of time series high and for which month is it low?

# The time series values have high values in the month of June and the lowest in the month of December.

In June ,it is a holiday season for many students and since everyone has some time off, they tend to buy certain items for their vacation. So it can be a good time to buy some gold so that they could wear it for their vacation. It can also be a good wedding season time due to the weather not being too cold. 

# 6. Show the plot for time series adjusted for seasonality. Overlay this with the line for actual time series? # Does seasonality have big fluctuations to the value of time series?

```{r}
plot(gold_time_series)
lines(seasadj(stl_decomp), main = "Seasonally adjusted plot", col="Green")
```

# Seasonality does not have alot of fluctuations with the orginal time series data. 

## Naive Method

# 1. Output
```{r}
gold_naive = naive(gold_time_series)
plot(gold_naive, main = "Naive Forecast")
```

# 2. Perform Residual Analysis for this technique.

# 3. Do a plot of residuals. What does the plot indicate?

```{r}
plot(gold_naive$residuals, main = "Naive forecast residuals plot for gold data")
```

# Inferences : (Pending)

# 4. Do a Histogram plot of residuals. What does the plot indicate?

```{r}
hist(gold_naive$residuals, main ='Histogram of Residuals')
```

# Inferences : The Hist plot shows us that the data is not normally distributed. It also tells us that the data                is biased.

# 5. Do a plot of fitted values vs. residuals. What does the plot indicate?

```{r}
cbind(Fitted = fitted(gold_naive),
      Residuals=residuals(gold_naive)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()

#Inferences: The graph shows Heteroscedasticity which means that the variance of the residuals is not constant across the range of fitted values. The values are also random.
```

# 6.  Do a plot of actual values vs. residuals. What does the plot indicate?

```{r}
cbind(Actual = gold_time_series,
      Residuals=residuals(gold_naive)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()

#Inferences : The Actual vs residuals plot is very similar to the fitted vs residuals. The graph shows Heteroscedasticity which means that the variance of the residuals is not constant across the range of fitted values. The values are also random.
```

# 7. Do an ACF plot of the residuals? What does this plot indicate?

```{r}
Acf(gold_naive$residuals, main = "ACF of Naive residuals")

# Inferences :  Values of the Acf have not crossed the confidence level meaning there is no trend in the residuals. 
```

## Print the 5 measures of accuracy for this forecasting technique

```{r}
naive_accuracy <- accuracy(gold_naive)
naive_accuracy
```

#### Forecast

# 1. Time series value for next year. Show table and plot

```{r}
forecast(gold_time_series)
plot(forecast(gold_time_series))
```

# 2. Summarize this forecasting technique

# How good is the accuracy? : The RMSE is pretty high for this forecast which means that we might not be able to proceed with it. 

# What does it predict the value of time series will be in one year? : 188.1347


## Simple Moving Averages

# 1. Plot the graph for time series
# 2. Show the Simple Moving average of order 3 on the plot above in Red
# 3. Show the Simple Moving average of order 6 on the plot above in Blue
# 4. Show the Simple Moving average of order 9 on the plot above in Green

```{r}
mavg_forecast3 = ma(gold_time_series,order=3)
mavg_forecast6 = ma(gold_time_series,order=6)
mavg_forecast9 = ma(gold_time_series,order=9) 
plot(gold_time_series, main = "Plot along with moving averages")
lines(mavg_forecast3, col="Red")
lines(mavg_forecast6, col="Blue")
lines(mavg_forecast9, col="Green")

#Inferences : MAV of 3 is a better overlapping forecast as ciompared to other ones. 
```

# 5. (Bonus) show the forecast of next 12 months using one of the simple average order that you feel works best for time series. 

```{r}
MA3_forecast <- forecast(mavg_forecast3, h = 12)
plot(MA3_forecast)

```

# 6. What are your observations of the plot as the moving average order goes up? (pending)



## Simple Smoothing

# 1. Perform a simple smoothing forecast for next 12 months for the time series.

```{r}
ses_data <- ses(gold_time_series)
plot(ses_data)
```


# 2. What is the value of alpha? What does that value signify? 

# Ans : 0.8983. Alpha specifies the coefficient for the level smoothing. Values near 1.0 mean that the latest value has more weight.

# 3. What is the value of initial state? 

#Ans : 121.59

# 4. What is the value of sigma? What does the sigma signify? : 

#Ans : 6.9394. Sigma defines the variance in the forecast predicted.

```{r}
summary(ses_data)
```


# Perform Residual Analysis for this technique

# Do a plot of residuals. What does the plot indicate?

```{r}
plot(ses_data$residuals, main = "Simple Smoothing forecast residuals plot")
```

# Inferences : The residuals look like they have seasonality. They do not have a mean at zero.


# Do a Histogram plot of residuals. What does the plot indicate?

```{r}
hist(ses_data$residuals, main ='Histogram of Residuals')
```

# Inferences : The histplot a normal distribution in an ideal case. The histogram does not appears to be normally distributed & their values do not have a mean zero. This means that the data is biased as the mean is not zero.


###### Do a plot of fitted values vs. residuals. What does the plot indicate?

```{r}
cbind(Fitted = fitted(ses_data),
      Residuals=residuals(ses_data)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()
```

#Inferences : The Actual vs residuals plot is very similar to the fitted vs residuals. The graph shows Heteroscedasticity which means that the variance of the residuals is not constant across the range of fitted values. The values are also random.


###### Do a plot of actual values vs. residuals. What does the plot indicate?

```{r}
cbind(Actual = gold_time_series,
      Residuals=residuals(ses_data)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()
```


#Inferences : The Actual vs residuals plot is very similar to the fitted vs residuals. The graph shows Heteroscedasticity which means that the variance of the residuals is not constant across the range of fitted values. The values are also random.



###### Do an ACF plot of the residuals? What does this plot indicate?

```{r}
Acf(ses_data$residuals, main = "ACF of Naive residuals")
```

#Inferences : The values have not crossed the confidence intervals & there is no seasonality. 


## Print the 5 measures of accuracy for this forecasting technique

```{r}
ses_accuracy <- accuracy(ses_data)
ses_accuracy
```


#### Forecast

###### Time series value for next year. Show table and plot

```{r}
forecast(ses_data)
plot(forecast(ses_data))
```

# Summarize this forecasting technique

#1. How good is the accuracy? : The ME, RMSE values are very high which tells us that this method may not be the the optimal one. 

#2. What does it predict the value of time series will be in one year? : 188.1548 (Sept 2024)

#3. Other observation : Simple smoothing forecast may not be a right way to forecast. We can consider more forecasting techniques and check if the error values are less than this one.



## HoltWinters

# 1. Perform Holt-Winters forecast for next 12 months for the time series.

```{r}
HW_forecast <- hw(gold_time_series, seasonal = "additive")
plot(forecast(HW_forecast))
attributes(HW_forecast)
hw_add <- forecast(HW_forecast)
hw_add$model
```


#Inferences : (Pending)


#1 What is the value of alpha? What does that value signify? 

#Ans : 0.9262 . Alpha specifies the coefficient for the level smoothing in Holtwinters.

#2. What is the value of beta? What does that value signify?

#Ans : 2e-04 . Beta specifies the coefficient for the trend smoothing in Holtwinters.

#3.What is the value of gamma? What does that value signify?

#Ans : 1e-04 Gamma specifies the coefficient for the trend smoothing in Holtwinters.


#4. What is the value of initial states for the level, trend and seasonality? What do these values signify?

#Ans : l = 121.6981 ,  b = 1.5747 , s = -5.2835 -4.7104 -3.8007 4.2063 4.4507 0.7968 4.4918 3.4419 -0.5184                                              -3.1621 1.166 -1.0784


#5.What is the value of sigma? What does the sigma signify?

#Ans : 7.1693 . Sigma defines the variance of the forecast values.


# Perform Residual Analysis for this technique.

# 1. Do a plot of residuals. What does the plot indicate?

```{r}
plot(hw_add$residuals, main = "Simple Smoothing forecast residuals plot")
```

#Inferences : The residuals do have a mean at zero for a lot of values as compared to previous ones. This looks like the best case. 

# Do a Histogram plot of residuals. What does the plot indicate?

```{r}
hist(hw_add$residuals, main ='Histogram of Residuals')
```
#Inferences : The data looks to be normally distributed. 


#. Do a plot of fitted values vs. residuals. What does the plot indicate?

```{r}
cbind(Fitted = fitted(hw_add),
      Residuals=residuals(hw_add)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()
```

#Infeences : The Fitted vs Residuals plot is random.

# Do a plot of actual values vs. residuals. What does the plot indicate?

```{r}
cbind(Actual = gold_time_series,
      Residuals=residuals(hw_add)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()
```

#Inferences : The Actual vs Residuals plot is random.


###### Do an ACF plot of the residuals? What does this plot indicate?

```{r}
Acf(hw_add$residuals, main = "ACF of Naive residuals")
```
#Inferences : From Acf plot, we can see that none of the values crossed the confidence levels.This shows us that the forecast is a good as compared to others. 


# Print the 5 measures of accuracy for this forecasting technique

```{r}
hw_accuracy <- accuracy(hw_add)
hw_accuracy
```


# Forecast

#1.Time series value for next year. Show table and plot

```{r}
forecast(hw_add)
plot(forecast(hw_add))
```

# Summarize this forecasting technique

# How good is the accuracy? : The ME & The RMSE values are lower than the previous forecasts. This is another reason why it is the best forecast.

# What does it predict the value of time series will be in one year? : 226.1218 (Nov 2025)


# Other observation : Holtwinters appears to be the best forecast considering all the previous forecast methods. We can still improve it if ARIMA works better. 


#Arima 

#1. Is Time Series data Stationary? How did you verify?
#2. How many differences are needed to make it stationary?

```{r}
ndiffs(gold_time_series)
```
#Inferences : this has a ndiff of 1 which means that the time series data is not stationary and will need one more difference to become stationary. It also means that the series shows a trend. 

#3. Is Seasonality component needed?

#Ans : yes, since it is non stationary, seasonality is needed.

#4. Plot the Time Series chart of the differenced series.
#5. Plot the ACF and PACF plot of the differenced series.

tsdisplay(gold_time_series)

#6. Based on the ACF and PACF, which are the possible ARIMA model possible?



#7. Show the AIC, BIC and Sigma^2 for the possible models?

#8. Based on the above AIC, BIC and Sigma^2 values, which model will you select?

#9. What is the final formula for ARIMA with the coefficients?

```{r}
auto_fit <- auto.arima(gold_time_series, trace = TRUE, stepwise = FALSE)
attributes(auto_fit)
plot(forecast(auto_fit, h = 1, level = c(99.5)))
```

#10. Perform Residual Analysis for this technique.

plot(auto_fit$residuals , main = "ARIMA residuals Plot")



#11 Do a plot of residuals. What does the plot indicate?


```{r}
plot(auto_fit$residuals, main = "ARIMA residuals plot")
```
#Inferences : The plot shows there are values around the mean of zero. 

#12 Do a Histogram plot of residuals. What does the plot indicate?

hist(auto_fit$residuals, main = "Histogram of residuals")

#Inferences : The hist plot shows us that there is not a normally distributed curve. 

#13 Do a plot of fitted values vs. residuals. What does the plot indicate?

```{r}
cbind(Fitted = fitted(ses_data),
      Residuals=residuals(ses_data)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()
```

#Inference : The values are random here. The residuals may not be constant. 

#14 Do a plot of actual values vs. residuals. What does the plot indicate?

```{r}
cbind(Actual = gold_time_series,
      Residuals=residuals(ses_data)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()

```

#Inference : The values are random here. The residuals may not be constant. 

#15 Do an ACF plot of the residuals? What does this plot indicate?

Acf(auto.fit$residuals)

#Inferences : There are no lag values crossing the confidence interval. 


#16. Print the 5 measures of accuracy for this forecasting technique.

```{r}
arima_accuracy <- accuracy(auto_fit)
arima_accuracy
```

#Inferences : The ME and RMSE are not better than the Holt winters one. So this is again 2nd best to HW.


##Forecast

# Next one year. Show table and plot

forecast(auto_fit, h = 12)
plot(forecast(auto_fit, h =12))

# Next two years. Show table and plot

forecast(auto_fit, h = 24)
plot(forecast(auto_fit, h =24))


#Summarize this forecasting technique

#How good is the accuracy?

#Inferences : the accuracy is 2nd best to Holt winters. 

#What does it predict time series will be in one year and next two years? 

# 188.75 in next 2 years and 188.75 in next 1 year. 

#Other observation

## Accuracy Summary

#### Show a table of all the forecast method above with their accuracy measures

```{r}
accuracy_table <- data.frame(
  Method = c("Naive", "Simple Smoothing", "Holt-Winters"),
  ME = c(naive_accuracy[1, "ME"], ses_accuracy[1, "ME"], hw_accuracy[1, "ME"]),
  MAE = c(naive_accuracy[1, "MAE"], ses_accuracy[1, "MAE"], hw_accuracy[1, "MAE"]),
  MASE = c(naive_accuracy[1, "MASE"], ses_accuracy[1, "MASE"], hw_accuracy[1, "MASE"]),
  RMSE = c(naive_accuracy[1, "RMSE"], ses_accuracy[1, "RMSE"], hw_accuracy[1, "RMSE"]),
  MAPE = c(naive_accuracy[1, "MAPE"], ses_accuracy[1, "MAPE"], hw_accuracy[1, "MAPE"])
)

print(accuracy_table)
```


##Conclusion


#1. Summarize your analysis of time series value over the time-period.

#The data does have trend and seasonality and we found the same with Acf plots. From the forecasting methods naive, simple smoothing, and HoltWinters, we can see that HoltWinters forecast gives us the best forecasting method in this case.
#Forecast is good because the error values are quiet low for HoltWinters.
#Residuals in HoltWinters appear to be random and the ACF values of residuals are inside the confidence intervals.  Naive is the worst since it does not have better forecasts as compared to the rest. 



#2. Based on your analysis and forecast above, do you think the value of the time series will increase, decrease or stay flat over the next year? How about next 2 years?

#the time series will increase over the next & next 2 years.
